% Cornell CS 2043: Unix Tools and Scripting
% Spring 2016
%
% http://cs2043-sp16.github.io/
%
% These slides have been created using the metroplis theme demo found here:
%
%     https://github.com/matze/mtheme.git
%
% I am a huge fan of the metropolis theme...thanks @matze!

\documentclass[11pt]{beamer}
\usetheme{metropolis}
\metroset{background=dark} % easier on the eyes, in my opinion
\setbeamertemplate{bibliography item}{\insertbiblabel}

% linking
\usepackage{hyperref}

% Nicer tables
\usepackage{booktabs}
\usepackage[scale=2]{ccicons}

% finer control over enumerations
% example:
%   \begin{enumerate}[(1)]
%     \item first
%     \item second
%     \item ...
%   \end{enumerate}
%
%   and also
%
%   \begin{enumerate}[a)]
%   \begin{enumerate}[i.]
%   \begin{enumerate}[$\mathbb{R}$\textsuperscript{\arabic{enumi}}]
%
% and so on
\usepackage{enumerate}

% sultry code listings
\usepackage{minted}
\BeforeBeginEnvironment{minted}{\begin{block}{}} % These surround minted code with beamer blocks
\AfterEndEnvironment{minted}{\end{block}}        % to accentuate them

%
% Colors thanks to various themes from Monokai: http://www.colourlovers.com/lover/Monokai
%
\definecolor{make-it-concrete}{rgb}{0.670588, 0.666667, 0.596078}      % (171, 170, 152) / 255
\definecolor{cheap-carpet}{rgb}{0.145098, 0.196078, 0.200000}% (37,50,51) / 255
\definecolor{hairball}{rgb}{0.403922, 0.411765, 0.380392}% (103,105,97) / 255

%
% General slide coloring
%
\setbeamercolor{frametitle}{% adjusts too-white colors from beamer-metropolis
  bg=hairball!66,
  fg=cheap-carpet
}
\setbeamercolor{normal text}{
  bg=mDarkTeal,   % note: defined by beamer-metropolis
  fg=hairball!44
}
%
% beamer-block shaping and colors
%
\setbeamertemplate{blocks}[rounded]
\setbeamercolor{block title}{
  bg=make-it-concrete!66,
  fg=cheap-carpet
}
\setbeamercolor{block body}{
  bg=cheap-carpet,
  fg=make-it-concrete!66
}

% Custom commands
\newcommand{\colbf}[1]{\textcolor{mLightBrown!77!black}{#1}}% bold with color; mLightBrown defined by beamer-metropolis
\newcommand{\tsub}{\textsubscript}                          % more compact
\newcommand{\tsup}{\textsuperscript}                        % " " " " " "
\def\wl{\par \vspace{\baselineskip}}                        % force newline; MANY thanks:
                                                            %   https://groups.google.com/d/msg/comp.text.tex/abxw5nTALzg/puXHnyx6pM8J

% Titles etc
\title{01 \-- Introduction}
\subtitle{CS 2043: Unix Tools and Scripting, Spring 2016 \cite{prevSemesters}}
\date{January 27th, 2016}
\author{Stephen McDowell}
\institute{Cornell University}

\begin{document}
\maketitle

% TOC
\begin{frame}{Table of contents}
  \setbeamertemplate{section in toc}[sections numbered]
  \tableofcontents[hideallsubsections]
\end{frame}

%
%%
%%%
%%%%
%%%%% section welcome
%%%%%
%%%%%
%%%%%
%%%%%
\section{Welcome}
\label{sec:welcome}

\begin{frame}{Redacted...}
  \begin{itemize}
    \item Due to technical complications on the first day, and reduced attendance due to Snowzilla,  I choose
          not to release the original slides.
    \item All we did was talk about course logistics, which you can find at
          \href{http://cs2043-sp16.github.io/}{\texttt{\colbf{http://cs2043-sp16.github.io/}}}.
    \item My original slides were greatly tailored to the words that would have been said.  Ultimately, all of
          the information presented can be obtained from the course website.
  \end{itemize}
\end{frame}

\begin{frame}{Redacted...}
  \href{http://www.cs.cornell.edu/courses/cs2043/2015sp/lectures/lecture01.pdf}{\colbf{Last year's intro lecture}}
  gives an excellent introduction to UNIX, the history, and the goals of this course. Note that some of the
  logistics of this year are quite different \-- anything written on the current semester's course website
  takes precedence.  A couple of notes about the differences:

  \begin{itemize}
    \item You must use a VM in this course if you are using Windows, Cygwin is not sufficient.
    \item I strongly discourage configuring a dual-boot \emph{at this time}.  We will talk about it more later...
    \item You do not have \texttt{ssh} access yet, I am waiting for enrollment to stabilize
    \item OSX: do \colbf{not} install Macports.  It is far inferior to \emph{brew}.  We will discuss \emph{brew}
          later in the course.
  \end{itemize}
\end{frame}

\begin{frame}{Redacted...}
  You are welcome to browse the other slides posted for last year, but do keep in mind that this is a different
  course and I will be covering things at a different level as well as the order will likely be switched.
\end{frame}

%%%%%
%%%%%
%%%%%
%%%%%
%%%%% section welcome
%%%%
%%%
%%
%

\begin{frame}[allowframebreaks]{References}
  \bibliography{lecture01/references}
  \bibliographystyle{abbrv}
\end{frame}

\end{document}
